Thesis: draft

### 2021-05-28 draft.tex
latest update

The stochastic Watts-Strogatz block model is used as a model that has both high clustering and community structure. We will check if the combination of both features displays any difference in the formation of echo chambers compared to having only a single feature (either community structure or high clustering).

@@

The models discussed in the previous sections are idealized and simplified. They may posses some of the characteristics of real-world networks, but they do not always describe real-world networks appropriately. Here, we will list some of the main characteristics of real-world networks, and more specifically of real-world social networks. A relevant question is whether there is a significant difference between off-line social networks and on-line social networks.\\

@@

We model opinion dynamics on social media platforms, with focus on Facebook. Each time we access Facebook we experience algorithmic personalization. The time-line or News Feed of each user on Facebook is the complex combination of posts from, among other, Facebook friends, companies and mass media. In 2013, Facebook reported that on average 1500 potential stories of friends and pages we follow or companies that are sponsored could be shown on the News Feed of each user \cite{facebook}. No user could ever process this amount of information, hence, Facebook selects around 300 of them, which then appear on your personalized News Feed. Their selection methods are based on (i) how often we interact with certain friends, pages or prominent figures; (ii) the amount of likes, shares and comments a post receives from our friends and others; (iii) how much we have interacted with certain types of posts before and (iv) whether we and other people on Facebook are hiding or reporting a certain post \cite{facebook}. \\
\newline
If we translate Facebook to a social network, we end up with a temporal network. Nodes represent the Facebook users, which become active and inactive over time. Undirected edges represent friendship ties on Facebook, that is, they link an user to his Facebook friends and the pages he or she follows. Each node then has an attribute that needs to mimic the Facebook News Feed and which is susceptible to the filtering algorithm.

@@

This thesis follows the opinion dynamics model that was designed by Perra and Rocha \cite{Perra2019}. We start with a network with $N = 10^3$ actors.  Initially, each actor is given one of two opinions, A or B. Each time step a random fraction equal to $p= 0.1$ of these actors becomes active, where $p$ is the constant activation probability. Then, all active actors update their opinion synchronously. This updating is done as follows. First, the active actor broadcasts its current opinion to all its neighbors. Then, it updates its own opinion according to the opinions on its time-line $R_i(t)$. This time-line contains only a selection of all the opinions received from its friends, which is the fundamental idea behind algorithmic personalization. Each actor has a hidden time-line $L_i(t)$ which contains the time-line of the actor at the last time it was active ($R_i(t_{\text{last}})$) plus all the post that its friends posted during $\interval[right open fence =), open right]{t_\text{last}}{t}$. If an actor becomes active, the filtering algorithm select $S = 20$ items from this hidden time-line $L_i(t)$ and shows them on the time-line $R_i(t)$ of the actor. Updating is done according to the probabilistic majority model: an actor who sees $20\%$ of the posts promoting opinion A and thus $80\%$ promoting opinion B will adopt opinion A with a chance $p_\text{A} = 0.2$ and opinion B with a chance $p_\text{B} = 0.8$. If another model is used, e.g. the majority model, this will be explicitly stated.
% Add pictorial representation of model??

@@

Perra and Rocha proposed different ways of selecting the $S$ items from the hidden time-line $L_i(t)$ \cite{Perra2019}. We will implement three of their algorithms: (i) REF filtering: $S$ random items from the hidden time-line $L_i(t)$ are selected and shown on the actual time-line $R_i(t)$; (ii) REC filtering: the $S$ most recent posts in the hidden time-line $L_i(t)$ are displayed on the actual time-line $R_i(t)$ and (iii) PR filtering: this way of filtering is based on semantic filtering explained in Section \ref{filter}; the posts on the hidden time-line $L_i(t)$ are ordered according to the current opinion of the actor (if the actor has opinion A, the hidden time-line will be ordered as AAAA...BBB), then the $S$ first items are selected and shown on the time-line $R_i(t)$.\\
\newline
The base-line algorithms are the REF and REC filtering algorithms. Filtering tailored to the user's preferences is accomplished with the PR filtering algorithm.

@@

If we want to introduce stubbornness in the model we will mainly use the resistance parameter, as explained in Section \ref{stubbBin}. Lets summarize this concept of stubbornness. Each actor is given a resistance parameter $r_i$, with $0 \leqslant r_i \leqslant 1$; $r_i = 0$ means not stubborn and $r_i = 1$ means completely stubborn. For all intermediate values of $r_i$, a random number $f$ between zero and one is drawn; if $f > r_i$, the actor updates its opinion, if $f \leqslant r_i$, the actor keeps its current opinion.

\begin{algorithm}[H]
\caption{Updating the opinions of the active actors at time $t$}
\label{alg:update}
\begin{algorithmic}[1]
	\State \(L_i \leftarrow \text{hidden time-line}\)
	\State \(R_i \leftarrow \text{actual time-line}\)
	\State \(S \leftarrow 20\) \Comment{Maximum length of time-line $R$}
	\State \(r\) \Comment{Stubbornness parameter of each actor, takes values between $0$ and $1$}
	\ForAll{active actor $i$}
		\State \(\text{op} \leftarrow \text{current opinion}\)
		\ForAll{neighbor $k$} \Comment{Add opinion of active actor $i$ to hidden time-line of each neighbor $k$}
			\State \(L_k.\textsc{add}\text{(op)}\)
		\EndFor
		\State \(f \leftarrow \text{rand(0, 1)}\) \Comment{Draw random number between $0$ and $1$}
		\If{$f > r$}
			\State \(\text{op} \leftarrow \Call{Update opinion}{L_i}\) \Comment{Update opinion of active actor $i$}
		\Else
			\State \(\text{op} \leftarrow \text{current opinion}\) \Comment{Do not update opinion of active actor $i$}
		\EndIf
	\EndFor

	\Function{Update opinion}{$L_i$}
		\State \(\textsc{Order}(L_i)\) \Comment{Order hidden time-line according to REF, REC or PR filtering}
		\If{$\text{len}(L_i) > S$}
			\State \(R_i \leftarrow L_i.\textsc{resize}(S)\) \Comment{Time-line $R_i$ = $S$ first elements of ordered hidden time-line}
		\EndIf
		\State \(\text{op} \leftarrow \textsc{probMaj}(R_i)\) \Comment{Update opinion according to probabilistic majority model}
	\State \Return op \Comment{Return updated opinion}
	\EndFunction
\end{algorithmic}
\end{algorithm}
\noindent
The pseudo-code for updating the opinions of the active actors in case of using a resistance parameter $r$ at time step $t$ is displayed in Algorithm \ref{alg:update} to make the model more comprehensive and understandable.

@@

\begin{figure}[H]
  \begin{subfigure}[b]{0.24\textwidth}
    \caption{}
  	\includegraphics[width=4cm, height=3cm]{drawing.png}
    \label{pr_real_PGP}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \caption{}
  	\includegraphics[width=4cm, height=3cm]{drawing1.png}
    \label{pr_real_PGP}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \caption{}
  	\includegraphics[width=4cm, height=3cm]{drawing2.png}
    \label{pr_real_PGP}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \caption{}
  	\includegraphics[width=4cm, height=3cm]{drawing3.png}
    \label{pr_real_PGP}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of stubborn actors for balanced starting conditions. The PR and REC filtering algorithms are compared. The PGP trust network is compared to corresponding network models.]{test.}
\label{echo_vs_fracRes_real_PGP}
\end{figure}

>>>>

The stochastic Watts-Strogatz block model is used as a model that has both high clustering and community structure. We will check if the combination of both features displays any difference in the formation of echo chambers compared to having only a single feature (either community structure or high clustering).\\
Fig. \ref{netRepr} visualizes the three main network models we used: (i) ER; (ii) WS and (iii) SBM.

\begin{figure}[H]
  \begin{subfigure}[b]{0.32\textwidth}
    \caption{}
    \includegraphics[width=5cm, height=5cm]{ER.png}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \caption{}
    \includegraphics[width=5cm, height=5cm]{WS.png}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \caption{}
    \includegraphics[width=5cm, height=5cm]{SBM.png}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Visualization of the three main network models used in this thesis; (i) ER; (ii) WS and (iii) SBM.]{Visualization of the three main network models used in this thesis. \textbf{(a)} ER; \textbf{(b)} WS; \textbf{(c)} SBM. The large number of triangles, indicating high clustering, are clearly visible for the WS model. The communities in the SBM are indicated with different colors. The definition of weak communities defines the community structure.}
\label{netRepr}
\end{figure}

@@

The models discussed in the previous sections are idealized and simplified. They may posses some of the characteristics of real-world networks, but they do not always describe real-world networks appropriately. Here, we will list some of the main characteristics of real-world networks, and more specifically of real-world social networks. A relevant question is whether there is a significant difference between off-line social networks and on-line social networks.

@@

We model opinion dynamics on social networks to replicate the opinion dynamics on social media platforms, with focus on Facebook. Each time we access Facebook we experience algorithmic personalization. The time-line or News Feed of each user on Facebook is the complex combination of posts from, among other, Facebook friends, companies and mass media. In 2013, Facebook reported that on average 1500 potential stories of friends and pages we follow or companies that are sponsored could be shown on the News Feed of each user \cite{facebook}. No user could ever process this amount of information, hence, Facebook selects around 300 of them, which then appear on your personalized News Feed. Their selection methods are based on (i) how often we interact with certain friends, pages or prominent figures; (ii) the amount of likes, shares and comments a post receives from our friends and others; (iii) how much we have interacted with certain types of posts before and (iv) whether we and other people on Facebook are hiding or reporting a certain post \cite{facebook}. \\
\newline
If we translate Facebook to a social network, we end up with a temporal network. Nodes represent the Facebook users, which become active and inactive over time. Undirected edges represent friendship ties on Facebook, that is, they link an user to his Facebook friends and the pages he or she follows. Each node has an attribute that needs to mimic the Facebook News Feed and which is susceptible to the filtering algorithm.

@@

This thesis follows the opinion dynamics model that was designed by Perra and Rocha \cite{Perra2019}. We start with a network with $N = 10^3$ actors.  Initially, each actor is given one of two opinions, A or B. Each time step a random fraction equal to $p= 0.1$ of these actors becomes active, where $p$ is the constant activation probability. Then, all active actors update their opinion synchronously. This updating is done as follows. First, the active actor broadcasts its current opinion to all its neighbors. Then, it updates its own opinion according to the opinions on its time-line $R_i(t)$. This time-line contains only a selection of all the opinions received from its friends, which is the fundamental idea behind algorithmic personalization. Each actor has a hidden time-line $L_i(t)$ which contains the time-line of the actor at the last time it was active ($R_i(t_{\text{last}})$) plus all the post that its friends posted during $\interval[right open fence =), open right]{t_\text{last}}{t}$. If an actor becomes active, the filtering algorithm select $S = 20$ items from this hidden time-line $L_i(t)$ and shows them on the time-line $R_i(t)$ of the actor. Updating is done according to the probabilistic majority model: an actor who sees $20\%$ of the posts promoting opinion A and thus $80\%$ promoting opinion B will adopt opinion A with a chance $p_\text{A} = 0.2$ and opinion B with a chance $p_\text{B} = 0.8$.

@@

Perra and Rocha proposed different ways of selecting the $S$ items from the hidden time-line $L_i(t)$ \cite{Perra2019}. We will implement three of their algorithms: (i) REF filtering: $S$ random items from the hidden time-line $L_i(t)$ are selected and shown on the actual time-line $R_i(t)$; (ii) REC filtering: the $S$ most recent posts in the hidden time-line $L_i(t)$ are displayed on the actual time-line $R_i(t)$ and (iii) PR filtering: this way of filtering is based on semantic filtering explained in Section \ref{filter}; the posts on the hidden time-line $L_i(t)$ are ordered according to the current opinion of the actor (if the actor has opinion A, the hidden time-line will be ordered as AAAA...BBB), then the $S$ first items are selected and shown on the time-line $R_i(t)$.\\
\newline
The base-line algorithms are the REF and REC filtering algorithms. Filtering tailored to the user's preferences is accomplished with the PR filtering algorithm.\\
\newline
Fig. \ref{op_dyn_schem} shows a schematic representation of the opinion dynamics model. The activation mechanism and filtering algorithms are represented.

\begin{figure}[H]
  \begin{subfigure}[b]{0.24\textwidth}
    \caption{}
  	\includegraphics[width=4cm, height=3.5cm]{drawing.png}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \caption{}
  	\includegraphics[width=4cm, height=3.5cm]{drawing1.png}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \caption{}
  	\includegraphics[width=4cm, height=3.5cm]{drawing2.png}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \caption{}
  	\includegraphics[width=4cm, height=3.5cm]{drawing3.png}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Schematic representation of the opinion dynamics model.]{Schematic representation of the opinion dynamics model. Focus lies on the central, colored actor. The grey bar represents its hidden time-line $L_i(t)$. The central actor is connected to six other actors. We assume that these are only connected to this central actor, so their opinion will not be changed until the central actor becomes active. \textbf{(a)} Initial situation: every actor is randomly assigned an opinion; the hidden time-line of the central actor is empty; \textbf{(b)} and \textbf{(c)} Actors become active (indicated in red) and push their opinion on the hidden time-line of the central actor; \textbf{(d)} The central actor becomes active: $S = 4$ items of its hidden time-line are selected and shown on its actual time-line. (i) REF: the hidden time-line is randomly shuffled; (ii) REC: the hidden time-line is ordered from most recent to oldest; (iii) PR: the hidden time-line is ordered according to the current opinion of the central actor. For each filtering algorithm the probability that the central actor changes its opinion is indicated. Only the first four posts are taken into account, the rest is deleted.}
\label{op_dyn_schem}
\end{figure}

@@

If we introduce stubbornness in the model we mainly use the resistance parameter, as explained in Section \ref{stubbBin}. Lets summarize this concept of stubbornness. Each actor is given a resistance parameter $r_i$, with $0 \leqslant r_i \leqslant 1$; $r_i = 0$ means not stubborn and $r_i = 1$ means completely stubborn. For all intermediate values of $r_i$, a random number $f$ between zero and one is drawn; if $f > r_i$, the actor updates its opinion, if $f \leqslant r_i$, the actor keeps its current opinion. \\
\newline
The pseudo-code for updating the opinions of the active actors in case of using a resistance parameter $r$ at time step $t$ is displayed in Algorithm \ref{alg:update}.

\begin{algorithm}[H]
\caption{Updating the opinions of the active actors at time $t$.}
\label{alg:update}
\begin{algorithmic}[1]
	\State \(L_i \leftarrow \text{hidden time-line}\)
	\State \(R_i \leftarrow \text{actual time-line}\)
	\State \(S \leftarrow 20\) \Comment{Maximum length of time-line $R$}
	\State \(r\) \Comment{Stubbornness parameter of each actor, takes values between $0$ and $1$}
	\ForAll{active actor $i$}
		\State \(\text{op} \leftarrow \text{current opinion}\)
		\ForAll{neighbor $k$} \Comment{Add opinion of active actor $i$ to hidden time-line of each neighbor $k$}
			\State \(L_k.\textsc{add}\text{(op)}\)
		\EndFor
		\State \(f \leftarrow \text{rand(0, 1)}\) \Comment{Draw random number between $0$ and $1$}
		\If{$f > r$}
			\State \(\text{op} \leftarrow \Call{Update opinion}{L_i}\) \Comment{Update opinion of active actor $i$}
		\Else
			\State \(\text{op} \leftarrow \text{current opinion}\) \Comment{Do not update opinion of active actor $i$}
		\EndIf
	\EndFor

	\Function{Update opinion}{$L_i$}
		\State \(\textsc{Order}(L_i)\) \Comment{Order hidden time-line according to REF, REC or PR filtering}
		\If{$\text{len}(L_i) > S$}
			\State \(R_i \leftarrow L_i.\textsc{resize}(S)\) \Comment{Time-line $R_i$ = $S$ first elements of ordered hidden time-line}
		\EndIf
		\State \(\text{op} \leftarrow \textsc{probMaj}(R_i)\) \Comment{Update opinion according to probabilistic majority model}
	\State \Return op \Comment{Return updated opinion}
	\EndFunction
\end{algorithmic}
\end{algorithm}

@@
